EDIpack: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-06-01 05_examples.tex
Updated Bibliobraphy with correct naming conventions <Author><Year><JournalAcronym>[a-z]

\subsection{Interacting Bernevig-Hughes-Zhang model (Fortran API, {\tt ed\_mode=nonsu2})}
In this section we focus on the DMFT solution of the interacting
Bernevig-Hughes-Zhang (BHZ).
As shown in Ref.~\cite{Amaricci2023PR}, at moderate
interactions this model admits an excitonic phase in which opposite
orbital electrons and holes across the band gap bind and form a coherent phase \cite{Knolle2017PRL,Jia2020,Varsano2020NN,Blason,Amaricci2023PR,Giuli2023PRB}

We consider a system of two-orbital electrons on a square
lattice interacting through a Hubbard-Kanamori term and describing
a quantum spin Hall insulator \cite{Kane2005PRL,Bernevig2006S,Konig2007S,Hasan2010RMPa,Hohenadler2011PRL,Amaricci2015PRL,Tang2017NP,Amaricci2023PR,Paoletti2024PR}.
We consider a suitable matrix basis in terms of the Dirac
matrices $\Gamma_{a\a}=\sigma_a\otimes \tau_\a$, where $\sigma_a$ and
$\tau_\a$ are Pauli matrices, respectively, in the spin and orbital
pseudo-spin space. The  model Hamiltonian reads
$$
H = \sum_{k}\psi_{k}^\dagger H(k)\psi_{k} + H_{\rm int},
$$
where $\psi_{k}=[c_{1\uparrow k}, c_{2\uparrow k},
c_{1\downarrow k}, c_{2\downarrow k} ]^T$ is the spinor collecting
annihilation operator $c_{a\sigma k}$ destroying an electron at
orbital $a=1,2$ with spin  $\sigma=\up,\dw$ and lattice momentum
$k$. The non-interacting part of the Hamiltonian is:
$$
H(k) = \left[M-2t(\cos{k_x}+\cos{k_y}) \right]\Gamma_{03} +
   \lambda\sin{k_x}\Gamma_{31} -   \lambda\sin{k_y}\Gamma_{02},
$$
where $M$ is the mass term, which plays the role of a crystal
field splitting among the orbitals. The presence of this term breaks
the symmetry in the orbital pseudo-spin channel.
The  interaction reads: 
$$
   H_{\rm int} = (U-J)\frac{\hat{N}(\hat{N}-1)}{2} - J\left( \frac{1}{4}\hat{N}^2 +
   \hat{S_z}^2 - 2 \hat{T_z}^2\right),
 $$
 where $\hat{N}=\tfrac{1}{2}\psi_i^\dagger \Gamma_{00}\psi_i$ is the
total density operator,
$\hat{S_z}=\tfrac{1}{2}\psi_i^\dagger \Gamma_{30}\psi_i$ is the total
spin polarization operator and $\hat{T_z}=\tfrac{1}{2}\psi_i^\dagger
\Gamma_{03}\psi_i$ is the orbital pseudo-spin polarization operator.
This form corresponds to the density-density part of the
Kanamori interaction. We neglect the pair-hopping and spin-flip purely
for numerical reasons \cite{Amaricci2022CPC}. 
%
In the non-interacting regime this model describes a
quantum spin Hall insulator (QSHI) for $M<4t$ and a trivial Band
Insulator (BI) for $M>4t$.
The transition point at $M=4t$ describes the formation of a gapless Dirac state at  $k=[0,0]$.

@@

\paragraph{Results}
To capture the possible symmetry breaking in any excitonic channel
driven by the local interaction, we consider the vector order
parameter
$\vec{E}=[E_0,E_x,E_y,E_z]$, where $E_a=\langle\psi^\dagger
\Gamma_{a1}\psi \rangle$ \cite{Budich2014,Kunes2014PRB,Kaneko2015JOPCS,Kunes2015JOPCM,Knolle2017PRL,Guerci2019PRM,Geffroy2019PRL,Mazza2020PRL,De-Palo2023PRB}.

The first component $E_0$ describes the singlet excitonic state,
whereas the remaining ones correspond to the triplet states with
different spin orientation \cite{Blason,Amaricci2023PR}.   
The analysis of the strong-coupling regime as well as impurity
excitonic susceptibility available in \NAME suggest the possible
instability to an in-plane spin-triplet exciton phase, i.e. with $E_x$
and/or $E_y$ different from zero, see \cite{Amaricci2023PR}.
Interestingly, this state breaks several symmetries including
time-reversal and spin SU(2), which protect the topological state.

@@

Unlike the static mean-field description \cite{Blason}, the
transition from the QSHI topological state to the Excitonic Insulator 
(EI) becomes of first-order upon including the local dynamical fluctuations \cite{Paoletti2024PR,BellomiaKMH} contained in DMFT. 
The EI continuously evolves into a Mott Insulator (MI)
for larger interaction (neglecting for simplicity any 
antiferromagnetic ordering that would naturally occur in this
regime).
The \NAME solution of this model nicely captures the two transitions which involve breaking of the spin symmetry group SU(2).

@@

The essential features of the ground state phase diagram have been established by intensive multi-method analyses \cite{Rachel2018ROPIP}.
For $\lambda_\mathrm{so}=0$ the model describes a Dirac semimetal, up to large repulsion, and magnetizes to an isotropic Néel state above a critical $U/t$ ratio\footnote{In contrast to the expectations for a generic half-filled bipartite lattice, for which the ground state is antiferromagnetic for any nonzero interaction, this property descends from the peculiar properties of Dirac cones: the vanishing density of states
at the Fermi level renders the logarithmic instability ineffective, even in the presence of perfect nesting \cite{SorellaTosatti92,Graphene_RMP}.}.
For a finite spin-orbit coupling $\lambda_\mathrm{so}\neq0$, the quantum spin-Hall state results increasingly stable against magnetic ordering, due to its symmetry-protected nontrivial topology. Furthermore, the long-range ordering
eventually stabilized at large $U/t$, is characterized by a lowered spin symmetry, due to the inherent coupling of spin and lattice degrees of freedom that favors in-plane easy-axis magnetization
\cite{3dXY_KMH}, as predicted by Hartree-Fock and $t$-$J$ perturbative expansions \cite{Rachel2010PRB} and
later confirmed by Auxiliary-Field Quantum Monte Carlo (AFQMC) 
simulations \cite{Hohenadler2013ROPIP}.

@@

We evaluate the total energy of both the AFM$_\perp$
and AFM$_\parallel$ calculations and compare them to establish which  is favored as the ground state of the system. In \figu{fig:KMenergy}(a)
we show results for the energy difference $E_\parallel-E_\perp$ across a wide range of $U/t$ and $\lambda_\mathrm{so}$ values. To assess the
degree of accuracy of our total energy calculation, we report in
\figu{fig:KMenergy}(b) a $\lambda_\mathrm{so}=0$ benchmark against
Hartree-Fock (simple static mean-field, but inherently variational), 
DMFT with NRG solver and lattice AFQMC data, as taken from 
Ref.~\cite{Racz2020}.

\begin{figure}
\hspace{1cm} (a) \hspace{6.5cm} (b)\\
    \includegraphics[width=0.47\linewidth,trim={0 0 0 5mm},clip]{figures/KMH_energy.pdf}\hfill
    \includegraphics[width=0.45\linewidth]{figures/benchEnergy_honey.pdf}\\
    \caption{In panel (a) we report the total energy difference between 
    the in-plane and the
    off-plane AFM solutions found with \NAME at $T=0$. Negative values
    signal a ground state with in-plane magnetization. The plateau at zero
    energy difference marks the topological phase of the model (a 
    paramagnetic quantum spin-Hall insulator), at all points except on the $\lambda_\mathrm{so}=0$ line, where the ground state is an interacting
    Dirac liquid, at weak coupling, or an isotropic antiferromagnet at strong coupling. In panel (b) we compare the total energies, across the 
    $\lambda_\mathrm{so}=0$ line, to published data for Hartree-Fock,
    DMFT-NRG and AFQMC solutions \cite{Racz2020}.}
    \label{fig:KMenergy}
\end{figure}

@@

In \figu{fig:KMflake} we report results for a nanosystem of 
$N_\mathrm{ineq}=54$ sites, solved with the bath parametrization 
described in \equ{eq:afmx_dmft_ansatz}, for the in-plane 
antiferromagnetic state.
Panel (a) shows the interaction dependence of the local magnetization,
computed from the impurity ground state as  
$ \ibra S_x \iket = \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket \equiv \ibra n_{i,\rightarrow} \iket - \ibra n_{i,\leftarrow} \iket$,
so that we can define the $S_x$ spin occupation numbers as
\begin{equation*}
    \ibra n_{i,\rightarrow} \iket \equiv \frac{\ibra n_{i,\up} + n_{i,\dw} \iket +
    \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket}{2},
    \qquad
    \ibra n_{i,\leftarrow} \iket \equiv \frac{\ibra n_{i,\up} + n_{i,\dw} \iket -
    \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket}{2},
\end{equation*}
which trivially satisfy the conservation of the local average charge $\ibra n_{i,\rightarrow} + n_{i,\leftarrow} \iket = \ibra n_{i,\up} + n_{i,\dw} \iket$.
Having a direct expression for $\ibra n_{i,\rightarrow} \iket$ and $\ibra n_{i,\leftarrow} \iket$
we can verify that the impurity reduced density matrix, computed by 
tracing the bath states in the \texttt{ed\_mode=nonsu2} representation 
(see \ref{sSecRDM}), once diagonalized, takes the familiar form
\cite{Zanardi2002,XiDai2013,WalshLetter2019} 
\begin{equation}
    \rho^\mathrm{imp} = 
     \begin{pmatrix}
            \left\ibra\left(1-n_{i,\rightarrow}\right)\left(1-n_{i,\leftarrow}\right)\right\iket   &    0          &       0       &   0 \\
                    0       &    \left\ibra n_{i,\rightarrow}\left(1-n_{i,\leftarrow}\right)\right\iket  &       0       &   0 \\
                   0       &       0   &    \left\ibra\left(1-n_{i,\rightarrow}\right)n_{i,\leftarrow}\right\iket      &   0 \\
                    0       &       0   &       0          &    \left\ibra n_{i,\rightarrow} n_{i,\leftarrow}\right\iket
        \end{pmatrix} 
\end{equation}
in the basis of the {\it natural} spin-orbitals
\cite{BellomiaPhD,BellomiaKMH,Bellomia_intracorr}, 
whose occupation numbers
are given by $n_{i,\rightarrow}$ and $n_{i,\leftarrow}$.
As  discussed in \cite{BellomiaPhD,BellomiaKMH,Bellomia_intracorr},
by defining the spin traces of $\rho^\mathrm{imp}$, as
\begin{equation}
    \rho^\mathrm{imp}_\sigma = 
    \Tr_{\bar{\sigma}} \left[\rho^\mathrm{imp}\right] =
    \begin{pmatrix}
        \ibra n_{i,\sigma} \iket & 0 \\
        0 & \ibra n_{i,\bar{\sigma}} \iket
    \end{pmatrix}
\end{equation}
we can directly measure local correlations by defining the mutual information
between the impurity spin-orbitals
$I(\rightarrow:\leftarrow) = 
    S(\rho^\mathrm{imp}_{\rightarrow}) + 
    S(\rho^\mathrm{imp}_{\leftarrow}) -
    S(\rho^\mathrm{imp})$,
where $S(\cdot)$ denotes the von Neumann entropy of its argument.

>>>>

\subsection{Interacting Bernevig-Hughes-Zhang model (Fortran API, {\tt ed\_mode=nonsu2})}
In this section we focus on the DMFT solution of the interacting
Bernevig-Hughes-Zhang (BHZ).
As shown in Ref.~\cite{Amaricci2023PRB}, at moderate
interactions this model admits an excitonic phase in which opposite
orbital electrons and holes across the band gap bind and form a coherent phase \cite{Knolle2017PRL,Jia2020,Varsano2020NN,Blason2020PRB,Amaricci2023PRB,Giuli2023PRB}

We consider a system of two-orbital electrons on a square
lattice interacting through a Hubbard-Kanamori term and describing
a quantum spin Hall insulator \cite{Kane2005PRL,Bernevig2006S,Konig2007S,Hasan2010RMP,Hohenadler2011PRL,Amaricci2015PRL,Tang2017NP,Amaricci2023PRB,Paoletti2024PRB}.
We consider a suitable matrix basis in terms of the Dirac
matrices $\Gamma_{a\a}=\sigma_a\otimes \tau_\a$, where $\sigma_a$ and
$\tau_\a$ are Pauli matrices, respectively, in the spin and orbital
pseudo-spin space. The  model Hamiltonian reads
$$
H = \sum_{k}\psi_{k}^\dagger H(k)\psi_{k} + H_{\rm int},
$$
where $\psi_{k}=[c_{1\uparrow k}, c_{2\uparrow k},
c_{1\downarrow k}, c_{2\downarrow k} ]^T$ is the spinor collecting
annihilation operator $c_{a\sigma k}$ destroying an electron at
orbital $a=1,2$ with spin  $\sigma=\up,\dw$ and lattice momentum
$k$. The non-interacting part of the Hamiltonian is:
$$
H(k) = \left[M-2t(\cos{k_x}+\cos{k_y}) \right]\Gamma_{03} +
   \lambda\sin{k_x}\Gamma_{31} -   \lambda\sin{k_y}\Gamma_{02},
$$
where $M$ is the mass term, which plays the role of a crystal
field splitting among the orbitals. The presence of this term breaks
the symmetry in the orbital pseudo-spin channel.
The  interaction reads: 
$$
   H_{\rm int} = (U-J)\frac{\hat{N}(\hat{N}-1)}{2} - J\left( \frac{1}{4}\hat{N}^2 +
   \hat{S_z}^2 - 2 \hat{T_z}^2\right),
 $$
 where $\hat{N}=\tfrac{1}{2}\psi_i^\dagger \Gamma_{00}\psi_i$ is the
total density operator,
$\hat{S_z}=\tfrac{1}{2}\psi_i^\dagger \Gamma_{30}\psi_i$ is the total
spin polarization operator and $\hat{T_z}=\tfrac{1}{2}\psi_i^\dagger
\Gamma_{03}\psi_i$ is the orbital pseudo-spin polarization operator.
This form corresponds to the density-density part of the
Kanamori interaction. We neglect the pair-hopping and spin-flip purely
for numerical reasons \cite{Amaricci2022CPC}. 
%
In the non-interacting regime this model describes a
quantum spin Hall insulator (QSHI) for $M<4t$ and a trivial Band
Insulator (BI) for $M>4t$.
The transition point at $M=4t$ describes the formation of a gapless Dirac state at  $k=[0,0]$.

@@

\paragraph{Results}
To capture the possible symmetry breaking in any excitonic channel
driven by the local interaction, we consider the vector order
parameter
$\vec{E}=[E_0,E_x,E_y,E_z]$, where $E_a=\langle\psi^\dagger
\Gamma_{a1}\psi \rangle$ \cite{Budich2014PRL,Kunes2014PRB,Kaneko2015JOPCS,Kunes2015JOPCM,Knolle2017PRL,Guerci2019PRM,Geffroy2019PRL,Mazza2020PRL,De-Palo2023PRB}.

The first component $E_0$ describes the singlet excitonic state,
whereas the remaining ones correspond to the triplet states with
different spin orientation \cite{Blason2020PRB,Amaricci2023PRB}.   
The analysis of the strong-coupling regime as well as impurity
excitonic susceptibility available in \NAME suggest the possible
instability to an in-plane spin-triplet exciton phase, i.e. with $E_x$
and/or $E_y$ different from zero, see \cite{Amaricci2023PRB}.
Interestingly, this state breaks several symmetries including
time-reversal and spin SU(2), which protect the topological state.

@@

Unlike the static mean-field description \cite{Blason2020PRB}, the
transition from the QSHI topological state to the Excitonic Insulator 
(EI) becomes of first-order upon including the local dynamical fluctuations \cite{Paoletti2024PRB,BellomiaKMH} contained in DMFT. 
The EI continuously evolves into a Mott Insulator (MI)
for larger interaction (neglecting for simplicity any 
antiferromagnetic ordering that would naturally occur in this
regime).
The \NAME solution of this model nicely captures the two transitions which involve breaking of the spin symmetry group SU(2).

@@

The essential features of the ground state phase diagram have been established by intensive multi-method analyses \cite{Rachel2018ROPIP}.
For $\lambda_\mathrm{so}=0$ the model describes a Dirac semimetal, up to large repulsion, and magnetizes to an isotropic Néel state above a critical $U/t$ ratio\footnote{In contrast to the expectations for a generic half-filled bipartite lattice, for which the ground state is antiferromagnetic for any nonzero interaction, this property descends from the peculiar properties of Dirac cones: the vanishing density of states
at the Fermi level renders the logarithmic instability ineffective, even in the presence of perfect nesting \cite{Sorella1992ELE,Castro-Neto2009RMP}.}.
For a finite spin-orbit coupling $\lambda_\mathrm{so}\neq0$, the quantum spin-Hall state results increasingly stable against magnetic ordering, due to its symmetry-protected nontrivial topology. Furthermore, the long-range ordering
eventually stabilized at large $U/t$, is characterized by a lowered spin symmetry, due to the inherent coupling of spin and lattice degrees of freedom that favors in-plane easy-axis magnetization
\cite{Griset2012PRB}, as predicted by Hartree-Fock and $t$-$J$ perturbative expansions \cite{Rachel2010PRB} and
later confirmed by Auxiliary-Field Quantum Monte Carlo (AFQMC) 
simulations \cite{Hohenadler2013ROPIP}.

@@

We evaluate the total energy of both the AFM$_\perp$
and AFM$_\parallel$ calculations and compare them to establish which  is favored as the ground state of the system. In \figu{fig:KMenergy}(a)
we show results for the energy difference $E_\parallel-E_\perp$ across a wide range of $U/t$ and $\lambda_\mathrm{so}$ values. To assess the
degree of accuracy of our total energy calculation, we report in
\figu{fig:KMenergy}(b) a $\lambda_\mathrm{so}=0$ benchmark against
Hartree-Fock (simple static mean-field, but inherently variational), 
DMFT with NRG solver and lattice AFQMC data, as taken from 
Ref.~\cite{Raczkowski2020PRB}.

\begin{figure}
\hspace{1cm} (a) \hspace{6.5cm} (b)\\
    \includegraphics[width=0.47\linewidth,trim={0 0 0 5mm},clip]{figures/KMH_energy.pdf}\hfill
    \includegraphics[width=0.45\linewidth]{figures/benchEnergy_honey.pdf}\\
    \caption{In panel (a) we report the total energy difference between 
    the in-plane and the
    off-plane AFM solutions found with \NAME at $T=0$. Negative values
    signal a ground state with in-plane magnetization. The plateau at zero
    energy difference marks the topological phase of the model (a 
    paramagnetic quantum spin-Hall insulator), at all points except on the $\lambda_\mathrm{so}=0$ line, where the ground state is an interacting
    Dirac liquid, at weak coupling, or an isotropic antiferromagnet at strong coupling. In panel (b) we compare the total energies, across the 
    $\lambda_\mathrm{so}=0$ line, to published data for Hartree-Fock,
    DMFT-NRG and AFQMC solutions~\cite{Raczkowski2020PRB}.}
    \label{fig:KMenergy}
\end{figure}

@@

In \figu{fig:KMflake} we report results for a nanosystem of 
$N_\mathrm{ineq}=54$ sites, solved with the bath parametrization 
described in \equ{eq:afmx_dmft_ansatz}, for the in-plane 
antiferromagnetic state.
Panel (a) shows the interaction dependence of the local magnetization,
computed from the impurity ground state as  
$ \ibra S_x \iket = \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket \equiv \ibra n_{i,\rightarrow} \iket - \ibra n_{i,\leftarrow} \iket$,
so that we can define the $S_x$ spin occupation numbers as
\begin{equation*}
    \ibra n_{i,\rightarrow} \iket \equiv \frac{\ibra n_{i,\up} + n_{i,\dw} \iket +
    \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket}{2},
    \qquad
    \ibra n_{i,\leftarrow} \iket \equiv \frac{\ibra n_{i,\up} + n_{i,\dw} \iket -
    \ibra c^\dagger_{i,\up} c_{i,\dw} + c^\dagger_{i,\dw} c_{i,\up} \iket}{2},
\end{equation*}
which trivially satisfy the conservation of the local average charge $\ibra n_{i,\rightarrow} + n_{i,\leftarrow} \iket = \ibra n_{i,\up} + n_{i,\dw} \iket$.
Having a direct expression for $\ibra n_{i,\rightarrow} \iket$ and $\ibra n_{i,\leftarrow} \iket$
we can verify that the impurity reduced density matrix, computed by 
tracing the bath states in the \texttt{ed\_mode=nonsu2} representation 
(see \ref{sSecRDM}), once diagonalized, takes the familiar form
\cite{Zanardi2002PRA,Su2013MPLB,Walsh2019PRL} 
\begin{equation}
    \rho^\mathrm{imp} = 
     \begin{pmatrix}
            \left\ibra\left(1-n_{i,\rightarrow}\right)\left(1-n_{i,\leftarrow}\right)\right\iket   &    0          &       0       &   0 \\
                    0       &    \left\ibra n_{i,\rightarrow}\left(1-n_{i,\leftarrow}\right)\right\iket  &       0       &   0 \\
                   0       &       0   &    \left\ibra\left(1-n_{i,\rightarrow}\right)n_{i,\leftarrow}\right\iket      &   0 \\
                    0       &       0   &       0          &    \left\ibra n_{i,\rightarrow} n_{i,\leftarrow}\right\iket
        \end{pmatrix} 
\end{equation}
in the basis of the {\it natural} spin-orbitals
\cite{BellomiaPhD,BellomiaKMH,Bellomia_intracorr}, 
whose occupation numbers
are given by $n_{i,\rightarrow}$ and $n_{i,\leftarrow}$.
As  discussed in \cite{BellomiaPhD,BellomiaKMH,Bellomia_intracorr},
by defining the spin traces of $\rho^\mathrm{imp}$, as
\begin{equation}
    \rho^\mathrm{imp}_\sigma = 
    \Tr_{\bar{\sigma}} \left[\rho^\mathrm{imp}\right] =
    \begin{pmatrix}
        \ibra n_{i,\sigma} \iket & 0 \\
        0 & \ibra n_{i,\bar{\sigma}} \iket
    \end{pmatrix}
\end{equation}
we can directly measure local correlations by defining the mutual information
between the impurity spin-orbitals
$I(\rightarrow:\leftarrow) = 
    S(\rho^\mathrm{imp}_{\rightarrow}) + 
    S(\rho^\mathrm{imp}_{\leftarrow}) -
    S(\rho^\mathrm{imp})$,
where $S(\cdot)$ denotes the von Neumann entropy of its argument.
